---
title: "Math 342W Final Project — Queens Apartment Price Prediction"
author: "Danielle Collins"
date: "May 26th 2026"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Packages & Setup

```{r}

## Core Libaries
library(tidyverse)
library(magrittr)

install.packages("stringr")
library(stringr)

## Modeling
install.packages("rJava")
install.packages("devtools")
devtools::install_github("kapelner/YARF/YARFJARs", ref = "dev")
devtools::install_github("kapelner/YARF/YARF", ref = "dev")


options(java.parameters = "-Xmx4000m")
install.packages("pacman")
library(pacman)
pacman::p_load(rJava)
library(rJava)
.jinit()

if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
pacman::p_load(YARF)

library(YARF)        # regression tree + random forest
library(rpart)       # fallback tree visualization
library(rpart.plot)

# Table formatting
library(knitr)
library(kableExtra)

# Missingness
library(missForest)  # for imputation (optional)
library(naniar)      # missingness summaries

set.seed(342)
```

---------------------------------------------

##Loading Our Data

```{r}
housing= read.csv("housing_data_2016_2017.csv", stringsAsFactors = FALSE)

dim (housing)
names(housing)

cat("Rows:", nrow(housing), "\n")
cat("Columns:", ncol(housing), "\n")

```
##Understanding Our Data 

```{r load-data}
##since coop_condo is our apartment type filter, we first check what the apartment type values are 

table(housing$coop_condo, useNA="always")

##Let's look at how our zip codes are stored
head(housing$full_address_or_zip_code,20)

##Check sale prices
summary(housing$sale_price)
head(housing$sale_price,20)

##Examine rest of data

table(housing$garage_exists,useNA="always")
##"yes", "Yes", "eys", "Underground", "UG",and "1" most likely all mean the same thing. "NA" probably means no garage

table(housing$cats_allowed, useNA="always")
##cats_allowed has "y" and "yes" as separate values

table(housing$dogs_allowed, useNA="always")
##dogs_allowed has "yes89"and "yes" as seperate values, we'll combine them as "yes"

table(housing$fuel_type, useNA="always")
##fuel_type has "other" and "Other" as separate values

table(housing$kitchen_type, useNA="always")
##kitchen_type has "combo" and "Combo" as separate values, along with: "eat in", "Eat In" and "eatin" and also "efficiemcy", "efficiency", "efficiency kitchen","efficiency kitchene" and "efficiency ktchen"

table(housing$dining_room_type, useNA="always")
table(housing$community_district_num, useNA = "always")

summary(housing$total_taxes)
summary(housing$parking_charges)
summary(housing$num_floors_in_building)
summary(housing$sq_footage)
summary(housing$num_bedrooms)
summary(housing$num_full_bathrooms)
summary(housing$approx_year_built)
summary(housing$walk_score)
summary(housing$maintenance_cost)
summary(housing$common_charges)
summary(housing$listing_price_to_nearest_1000)
summary(housing$num_floors_in_building)
summary(housing$num_half_bathrooms)
summary(housing$parking_charges)
summary(housing$pct_tax_deductibl)
summary(housing$total_taxes)
```
As can be seen from above, we have a lot of cleaning to do in our dataset before we can perform any modeling on it. Not only do we have missing values, but many features contain typos and duplicate values.

##Cleaning Our Data

```{r}
## our zip codes are stuck in address string, so we need to extract them

housing$zipcode= as.integer(stringr::str_extract(housing$full_address_or_zip_code, "\\d{5}"))
## taking anywhere we see 5 digits in a row

##Change sale price to numeric from string
housing$sale_price= as.numeric(gsub("[$,\\s]", "", housing$sale_price))
##gsub finds every match of a pattern in a string and replaces it- finding "$" and "," and removing it then turning final result into a number

##Change listing price to numeric from string
housing$listing_price_to_nearest_1000= as.numeric(gsub("[$,\\s]", "", housing$listing_price_to_nearest_1000))
##gsub finds every match of a pattern in a string and replaces it- finding "$" and "," and removing it then turning final result into a number

##Change total taxes to numeric from string
housing$total_taxes <- as.numeric(gsub("[$,\\s]", "", housing$total_taxes))

##Change parking charges to numeric from string
housing$parking_charges <- as.numeric(gsub("[$,\\s]", "", housing$parking_charges))

##Change maintenance cost to numeric
housing$maintenance_cost= as.numeric(gsub("[$,\\s]", "", housing$maintenance_cost))

##Change common charges to numeric
housing$common_charges= as.numeric(gsub("[$,\\s]", "", housing$common_charges))

##Now we have to deal with our typos and duplicates in feature values

## Standardizing cats_allowed
housing$cats_allowed = ifelse(housing$cats_allowed=="y","yes", housing$cats_allowed)
housing$cats_allowed <- tolower(housing$cats_allowed)
##Replaces y with yes, and for entries with "Yes" or "No", changes to standard "yes" or "no"

## Standardizing dogs_allowed 
housing$dogs_allowed = ifelse(housing$dogs_allowed=="yes89","yes", housing$dogs_allowed)
housing$dogs_allowed= tolower(housing$dogs_allowed)

##Standardizing fuel_type
housing$fuel_type <- tolower(housing$fuel_type)

##Standardizing kitchen type
housing$kitchen_type= tolower(housing$kitchen_type)
housing$kitchen_type= ifelse(housing$kitchen_type %in% c("efficiemcy", "efficiency kitchen", "efficiency kitchene", "efficiency ktchen", "efficiency"), "efficiency", housing$kitchen_type)
housing$kitchen_type=ifelse(housing$kitchen_type=="1955", NA, housing$kitchen_type)
housing$kitchen_type= ifelse(housing$kitchen_type %in% c("combo", "Combo"), "combo", housing$kitchen_type)
housing$kitchen_type= ifelse(housing$kitchen_type== "eatin", "eat in", housing$kitchen_type)

##Standardizing dining_room_type

housing$dining_room_type = as.character(housing$dining_room_type)
housing$dining_room_type <- ifelse(housing$dining_room_type == "dining area","other", housing$dining_room_type)
##Changes the 2 entries that say "dining area" to "other"

##Standardizing garage exists
housing$garage_exists= tolower(housing$garage_exists)
housing$garage_exists= ifelse(housing$garage_exists %in% c("yes", "eys", "ug", "underground", "1"), "yes", housing$garage_exists)
housing$garage_exists= ifelse(is.na(housing$garage_exists), "no", housing$garage_exists)


summary(housing$sale_price)
summary(housing$maintenance_cost)
summary(housing$common_charges)
summary(housing$listing_price_to_nearest_1000)
summary(housing$total_taxes)
summary(housing$parking_charges)
summary(housing$num_floors_in_building)
table(housing$cats_allowed, useNA= "always")
table(housing$dogs_allowed, useNA= "always")
table(housing$fuel_type, useNA= "always")
table(housing$kitchen_type, useNA= "always")
table(housing$dining_room_type, useNA="always")
table(housing$garage_exists, useNA= "always")
table(housing$zipcode, useNA= "always")
table(housing$community_district_num, useNA = "always")

```

Now pay close attention to the zip codes. Upon closer inspection, we see that there are a few zip codes that do not belong to the Queens Borough. As a matter of fact, they're not even valid zip codes! Also, there are 15 addresses in which we can not find any zip codes for. Hence, our focus below will be on filtering our dataset to extract only those rows/observations with correct zip codes.

```{r}
valid_zipcodes= c(11361, 11362, 11363, 11364, 11354, 11355, 11356, 11357, 11358, 11359, 11360,11365, 11366, 11367,11412, 11423, 11432, 11433, 11434, 11435, 11436, 11101, 11102, 11103, 11104, 11105, 11106,11374, 11375, 11379, 11385,11004, 11005, 11411, 11413, 11422, 11426, 11427, 11428, 11429,11414, 11415, 11416, 11417, 11418, 11419, 11420, 11421,11368, 11369, 11370, 11372, 11373, 11377, 11378)


##Formatting dataset as how project specified
housing= housing %>% filter(zipcode %in% valid_zipcodes, coop_condo %in% c("co-op", "condo"), !is.na(sale_price), sale_price<=1000000, sale_price>0)


## Now lets check if we have any duplicate addresses in our filtered dataset (same address listed with same price, not referring to multiple units in a building)
dup_addresses <- housing$full_address_or_zip_code[
  duplicated(housing$full_address_or_zip_code)
]
cat("Number of duplicate addresses:", length(dup_addresses), "\n")

## View duplicates with sale price and date
result = housing[housing$full_address_or_zip_code %in% dup_addresses,
        c("full_address_or_zip_code", "sale_price", "date_of_sale")]
result= result[order(result$full_address_or_zip_code), ]
print(result)

## Remove true duplicates (same address, same price, same date)
housing= housing[!duplicated(
  paste(housing$full_address_or_zip_code, housing$sale_price, housing$date_of_sale)
), ]

##reset row numbers
rownames(housing) <- NULL

cat("Rows after removing duplicates:", nrow(housing), "\n")

dim(housing)
summary(housing$sale_price)

```

##Featurization
Now we have much less observations to work with, 518 to be exact, and also 57 features. As we saw in the beginning of this project, the number 57 comes from the MTurk metadata columns we have not gotten rid of yet. Also, we now need to officially select our desired features for our model. In the section below, we will focus addressing these matters collectively for a more meaningful dataset.

```{r}

housing= housing %>% mutate(age_at_sale= 2016- approx_year_built,
      ##converting yes/no features into binary and categorizing others
      garage_exists=ifelse(garage_exists=="yes",1,0),
                            
      is_coop=ifelse(coop_condo== "co-op",1,0),

      cats_allowed= ifelse(cats_allowed=="yes",1,0), 

      dogs_allowed = ifelse(dogs_allowed == "yes", 1, 0) ,

      kitchen_type = as.factor(kitchen_type), 

      dining_room_type = as.factor(dining_room_type),

      fuel_type = as.factor(fuel_type)

)

housing$community_district_num= as.factor(housing$community_district_num)

##Now, in addition to the features provided by our dataset, to achieve the most accurate model possible, we will introduce some external features (i.e. features related to our dataset but not currently in it). The values were extracted via national data collection during the time period of 2016-1017 relative to each zip code.

##Mean Income by Zipcode
mean_income_by_zip= data.frame(zips=c( 11004, 11101, 11102, 11103, 11104, 11105, 11106, 11109,
    11354, 11355, 11356, 11357, 11358, 11360, 11361, 11362,
    11363, 11364, 11365, 11366, 11367, 11368, 11369, 11370,
    11372, 11373, 11374, 11375, 11377, 11378, 11379, 11385,
    11411, 11412, 11413, 11414, 11415, 11416, 11417, 11418,
    11419, 11420, 11421, 11422, 11423, 11426, 11427, 11428,
    11429, 11432, 11433, 11434, 11435, 11436, 11691, 11692,
    11693, 11694, 11697
  ),
  mean_income = c(
    91500, 67000, 62000, 64000, 60000, 80000, 58500, 125000,
    53000, 50000, 76000, 98500, 85000, 88500, 86000, 100000,
    110000, 82000, 73000, 78000, 68000, 41500, 49500, 61000,
    55000, 48000, 66000, 80000, 59000, 73000, 85000, 57000,
    78000, 74000, 70000, 85000, 66000, 57000, 59000, 55000,
    53000, 59000, 57000, 78000, 67000, 90000, 78000, 76000,
    74000, 53000, 46000, 48000, 44000, 53000, 40000, 47000,
    69000, 86000, 122500
  )
  )

housing$mean_income= mean_income_by_zip$mean_income[
  match(housing$zipcode, mean_income_by_zip$zips)]

## Check it worked
sum(is.na(housing$mean_income))
summary(housing$mean_income)


# Grouping community districts- Why?
## Several districts have only 1 observation — districts 3, 4, 8, 16, 19, 20, 23 all have just 1 row each. These will randomly end up only in the test fold in later sections, resulting in errors.

housing$community_district_num <- as.character(housing$community_district_num)
housing$community_district_num <- ifelse(
  housing$community_district_num %in% c("24", "25", "26", "27", "28", "29", "30"),
  housing$community_district_num,
  "Other"
)
housing$community_district_num <- as.factor(housing$community_district_num)

table(housing$community_district_num)


##Crime Ranking by Zipcode
crime_by_zip <- data.frame(
  zips = c(
    11004, 11101, 11102, 11103, 11104, 11105, 11106, 11109,
    11354, 11355, 11356, 11357, 11358, 11360, 11361, 11362,
    11363, 11364, 11365, 11366, 11367, 11368, 11369, 11370,
    11372, 11373, 11374, 11375, 11377, 11378, 11379, 11385,
    11411, 11412, 11413, 11414, 11415, 11416, 11417, 11418,
    11419, 11420, 11421, 11422, 11423, 11426, 11427, 11428,
    11429, 11432, 11433, 11434, 11435, 11436, 11691, 11692,
    11693, 11694, 11697
  ),
  crime_rate = factor(c(
    "Very Safe", "High", "Moderate", "Moderate", "Moderate", "Safe", "Moderate", "Safe",
    "Moderate", "Moderate", "Safe", "Very Safe", "Very Safe", "Very Safe", "Very Safe", "Very Safe",
    "Extremely Safe", "Very Safe", "Safe", "Safe", "Safe", "High", "Moderate-High", "Moderate-High",
    "Moderate-High", "High", "Safe", "Safe", "Moderate", "Safe", "Very Safe", "Moderate-High",
    "Safe", "Safe", "Moderate", "Safe", "Safe", "Moderate", "Moderate", "Moderate",
    "Moderate", "Moderate", "Moderate", "Safe", "Moderate", "Very Safe", "Safe", "Safe",
    "Safe", "Moderate-High", "High", "High", "High", "Moderate-High", "High", "Moderate-High",
    "Moderate", "Safe", "Extremely Safe"
  ), levels = c("Extremely Safe", "Very Safe", "Safe", "Moderate", "Moderate-High", "High"))
)

housing$crime_rate= crime_by_zip$crime_rate[
  match(housing$zipcode, crime_by_zip$zips)]

sum(is.na(housing$crime_rate))
table(housing$crime_rate)

housing$neighborhood <- ifelse(housing$zipcode %in% c(11361, 11362, 11363, 11364), "Northeast Queens",
                        ifelse(housing$zipcode %in% c(11354, 11355, 11356, 11357, 11358, 11359, 11360), "North Queens",
                        ifelse(housing$zipcode %in% c(11365, 11366, 11367), "Central Queens",
                        ifelse(housing$zipcode %in% c(11412, 11423, 11432, 11433, 11434, 11435, 11436), "Jamaica",
                        ifelse(housing$zipcode %in% c(11101, 11102, 11103, 11104, 11105, 11106), "Northwest Queens",
                        ifelse(housing$zipcode %in% c(11374, 11375, 11379, 11385), "West Central Queens",
                        ifelse(housing$zipcode %in% c(11004, 11005, 11411, 11413, 11422, 11426, 11427, 11428, 11429), "Southeast Queens",
                        ifelse(housing$zipcode %in% c(11414, 11415, 11416, 11417, 11418, 11419, 11420, 11421), "Southwest Queens",
                        ifelse(housing$zipcode %in% c(11368, 11369, 11370, 11372, 11373, 11377, 11378), "West Queens", NA)))))))))

housing$neighborhood <- as.factor(housing$neighborhood)


table(housing$neighborhood)
sum(is.na(housing$neighborhood))

##Monthly Inflation Rate

## Extract month and year from date_of_sale
housing$sale_month <- as.integer(format(as.Date(housing$date_of_sale, "%m/%d/%Y"), "%m"))
housing$sale_year  <- as.integer(format(as.Date(housing$date_of_sale, "%m/%d/%Y"), "%Y"))

## US Monthly CPI (All Urban Consumers, 2016-2017)
cpi_data = data.frame(
  year  = c(2016, 2016, 2016, 2016, 2016, 2016, 2016, 2016, 2016, 2016, 2016, 2017, 2017),
  month = c(2,    3,    4,    5,    6,    7,    8,    9,    10,   11,   12,   1,    2),
  cpi   = c(236.9, 238.1, 239.3, 240.2, 241.0, 240.6, 240.8, 241.4, 241.7, 241.4, 241.4, 242.8, 243.6)
)

housing$cpi= cpi_data$cpi[
  match(paste(housing$sale_year, housing$sale_month), 
        paste(cpi_data$year, cpi_data$month))
]

sum(is.na(housing$cpi))
summary(housing$cpi)

## ---- Missingness Indicators ----
## Added before imputation so we capture which values were truly missing
## sq_footage and common_charges have very high missingness (60% and 75%), may be informative 

housing$sq_footage_missing <- as.integer(is.na(housing$sq_footage))
housing$common_charges_missing <- as.integer(is.na(housing$common_charges))
housing$maintenance_cost_missing <- as.integer(is.na(housing$maintenance_cost))
housing$num_floors_missing <- as.integer(is.na(housing$num_floors_in_building))

cat("sq_footage missing:", sum(housing$sq_footage_missing), "\n")
cat("common_charges missing:", sum(housing$common_charges_missing), "\n")
cat("maintenance_cost missing:", sum(housing$maintenance_cost_missing), "\n")
cat("num_floors missing:", sum(housing$num_floors_missing), "\n")

##now selecting our total final features

features = c(
  "sq_footage", 
  "age_at_sale",
  "num_bedrooms",
  "num_full_bathrooms",
  "num_total_rooms", 
  "is_coop", 
  "garage_exists",
  "cats_allowed", 
  "dogs_allowed",
  "kitchen_type", 
  "dining_room_type", 
  "fuel_type", 
  "maintenance_cost",
  "common_charges", 
  "walk_score",
  "mean_income", 
  "crime_rate",
  "neighborhood",
  "cpi",
  "num_floors_in_building", "community_district_num",
  "sq_footage_missing", "common_charges_missing", 
  "maintenance_cost_missing", "num_floors_missing"
  )

##Of note listing price was considered to be used as a feature originally to help estimate sale price as an approximation however:

## Check how many rows have both
both= sum(!is.na(housing$sale_price) & !is.na(housing$listing_price_to_nearest_1000))
cat("Rows with both sale price and listing price:", both, "\n")
##Therefore we will instead drop listing_price as a feature as there is no overlap between the two variables, every row either has a listing price or a sale price, never both. Since we need sale price as our outcome variable, we are restricted to the 521 rows that have it, and all of those rows have no listing price whatsoever. With 100% missingness in listing price across our modeling dataset, there is no information to impute from and no way to learn the relationship between listing price and sale price. We therefore drop it as a feature.


cat("Number of features:", length(features), "\n")
##Total features
dim(housing[,features])

##Our matrix has 521 observations(rows),18 features (columns)
summary(housing[,features])

"mean_income" %in% names(housing)
names(housing)

table(housing$neighborhood)
sum(is.na(housing$neighborhood))

##A few things to note: sq_footage and common_charges have 312 and 390 NAs respectively (over half of our observations), maintenance_cost has 140 NAs, kitchen_type- 7 NAs, dining_room_type- 115 NAs, age_at_sale- 6 NAs

```

##Imputation

That missingness we noted in the previous section is now our main point of focus in this one."Imputation" refers to the process of "filling in" missing values in a dataset with estimated or predicted values. Its purpose: To ensure completeness and improve our analysis' accuracy.

Usually, we would make use of the column mean as our estimated value. However, instead we pivot to using the column median due to cases such as the following: sq_footage has a max of 6215 which is much higher than its mean of 958, meaning that outlier would greatly pull our mean up and give us an unrealistic imputed value for square footage. 

For categorical features, we will use the mode as our estimated value.

```{r}

##Imputation via missForest 
library(missForest)

## Select only our feature columns for imputation
housing_for_imputation <- housing[, features]

## Run missForest
imputed <- missForest(housing_for_imputation)

## Replace feature columns with imputed values
housing[, features] <- imputed$ximp

## Check OOB imputation error
cat("missForest OOB imputation error:\n")
print(imputed$OOBerror)

## Confirm no NAs remain
cat("NAs remaining after missForest:", sum(is.na(housing[, features])), "\n")

```
##K-Fold Cross Validation

```{r}

## Final cross validation with all three models using tuned parameters
# mtry = 11 and num_trees = 500 were determined via tuning (see tuning section below)

set.seed(342)

n = nrow(housing)
k = 5
fold_idx <- sample(rep(1:k, length.out = n))

cat("n:", n, "\n")
cat("Fold sizes:\n")
print(table(fold_idx))

ols_rmse_folds  = c()
ols_r2_folds    = c()
tree_rmse_folds = c()
tree_r2_folds   = c()
rf_rmse_folds   = c()
rf_r2_folds     = c()

for (fold in 1:k) {
  
  test_idx  = which(fold_idx == fold)
  train_idx = which(fold_idx != fold)
  
  X_train = housing[train_idx, features]
  y_train = housing$sale_price[train_idx]
  X_test  = housing[test_idx, features]
  y_test  = housing$sale_price[test_idx]
  
  # --- OLS ---
  ols_model = lm(sale_price ~ ., data = cbind(X_train, sale_price = y_train))
  ols_preds = predict(ols_model, X_test)
  ols_rmse_folds[fold] = sqrt(mean((y_test - ols_preds)^2))
  ols_r2_folds[fold]   = 1 - sum((y_test - ols_preds)^2) / sum((y_test - mean(y_test))^2)
  
  # --- Regression Tree ---
  tree_model = YARFCART(X_train, y_train)
  tree_preds = predict(tree_model, X_test)
  tree_rmse_folds[fold] = sqrt(mean((y_test - tree_preds)^2))
  tree_r2_folds[fold] = 1 - sum((y_test - tree_preds)^2) / sum((y_test - mean(y_test))^2)
  
  # --- Random Forest (tuned) ---
  rf_model = YARF(X_train, y_train, num_trees = 500, mtry = 17)
  rf_preds = predict(rf_model, X_test)
  rf_rmse_folds[fold] = sqrt(mean((y_test - rf_preds)^2))
  rf_r2_folds[fold] = 1 - sum((y_test - rf_preds)^2) / sum((y_test - mean(y_test))^2)
  
  cat("Fold", fold, "complete\n")
}

cat("\n--- 5-Fold Cross Validation Results ---\n")
cat("OLS  RMSE:", round(mean(ols_rmse_folds), 2),  "R²:", round(mean(ols_r2_folds), 4), "\n")
cat("Tree RMSE:", round(mean(tree_rmse_folds), 2), "R²:", round(mean(tree_r2_folds), 4), "\n")
cat("RF   RMSE:", round(mean(rf_rmse_folds), 2),   "R²:", round(mean(rf_r2_folds),  4), "\n")


mean_sale_price = mean(housing$sale_price)

cat("\n--- Normalized RMSE (as proportion of mean sale price) ---\n")
cat("Mean sale price:", round(mean_sale_price, 2), "\n")
cat("OLS  Normalized RMSE:", round(mean(ols_rmse_folds) / mean_sale_price, 4), "\n")
cat("Tree Normalized RMSE:", round(mean(tree_rmse_folds) / mean_sale_price, 4), "\n")
cat("RF   Normalized RMSE:", round(mean(rf_rmse_folds) / mean_sale_price, 4), "\n")

```
## Tuning section for mtry and ntrees
```{r}
##Mtry
n = nrow(housing)
k = 5
fold_idx <- sample(rep(1:k, length.out = n))

cat("n:", n, "\n")
cat("Fold sizes:\n")
print(table(fold_idx))

mtry_values = c(9,11, 13, 15, 17, 19)
mtry_rmse = c()

for (m in mtry_values) {
  rmse_folds =  c()
  
  for (fold in 1:k) {
    test_idx  = which(fold_idx == fold)
    train_idx = which(fold_idx != fold)
    
    X_train = housing[train_idx, features]
    y_train = housing$sale_price[train_idx]
    X_test  = housing[test_idx, features]
    y_test  = housing$sale_price[test_idx]
    
    rf = YARF(X_train, y_train, num_trees = 500, mtry = m)
    preds = predict(rf, X_test)
    rmse_folds[fold] = sqrt(mean((y_test - preds)^2))
  }
  
  mtry_rmse = c(mtry_rmse, mean(rmse_folds))
  cat("mtry =", m, "RMSE:", round(mean(rmse_folds), 2), "\n")
}

optimal_mtry =  mtry_values[which.min(mtry_rmse)]
cat("\nBest mtry:", optimal_mtry, "\n")

##Ntrees

ntree_values = c(250, 500, 750, 1000)
ntree_rmse <- c()

for (nt in ntree_values) {
  rmse_folds = c()
  
  for (fold in 1:k) {
    test_idx  = which(fold_idx == fold)
    train_idx = which(fold_idx != fold)
    
    X_train = housing[train_idx, features]
    y_train = housing$sale_price[train_idx]
    X_test  = housing[test_idx, features]
    y_test  = housing$sale_price[test_idx]
    
    rf = YARF(X_train, y_train, num_trees = nt, mtry = optimal_mtry)
    preds = predict(rf, X_test)
    rmse_folds[fold] = sqrt(mean((y_test - preds)^2))
  }
  
  ntree_rmse = c(ntree_rmse, mean(rmse_folds))
  cat("num_trees =", nt, "RMSE:", round(mean(rmse_folds), 2), "\n")
}

optimal_ntrees = ntree_values[which.min(ntree_rmse)]
cat("\nBest num_trees:", optimal_ntrees, "\n")
```
##Final Fit on Dataset
```{r}
X_full = housing[,features]
y_full = housing$sale_price

##Final OLS
final_ols= lm(sale_price~.,data=cbind(X_full,sale_price=y_full))
summary(final_ols)

##Final Regression Tree
final_tree = YARFCART(X_full, y_full)

##Final Random Forest
final_rf= YARF(X_full, y_full, num_trees=500, mtry=11, calculate_oob_error = TRUE)

cat("RF OOB RMSE:", round(final_rf$rmse_oob, 2), "\n")
cat("RF OOB R²:  ", round(final_rf$pseudo_rsq_oob, 4), "\n")
```

##Visualizing Regression Tree for Interpretation
```{r}
illustrate_trees(final_tree, max_depth = 4, open_file = TRUE)

tree_rpart = rpart(sale_price ~ ., data = cbind(X_full, sale_price = y_full), control = rpart.control(maxdepth = 6))

var_imp = tree_rpart$variable.importance
var_imp_df = data.frame(
  Feature = names(var_imp),
  Importance = round(var_imp, 2)
) %>% arrange(desc(Importance))

kable(head(var_imp_df, 10),
      caption = "Top 10 Features by Variable Importance (Regression Tree)") %>%
  kable_styling(font_size = 9)

```

### Variable Importance for Random Forest

```{r}

var_counts = query_variable_counts(final_rf)
var_counts_df = data.frame(feature = names(var_counts), count = as.numeric(var_counts)) %>% arrange(desc(count))

## Bar Plot
barplot(head(var_counts_df$count, 10),
  names.arg = head(var_counts_df$Feature, 10),
  las = 2,
  main = "Random Forest Variable Importance (Split Count)",
  ylab = "Number of Splits",
  col = "steelblue",
  cex.names = 0.7
)

## Table
kable(head(var_counts_df, 10),
      caption = "Top 10 Features by Split Count (Random Forest)") %>% kable_styling(font_size = 9)


```

##Final Performance Results
```{r}
results = data.frame(
  Model = c("Regression Tree", "OLS Linear", "Random Forest"),
  `In-Sample R2` = c(
    round(1 - sum((y_full - predict(final_tree, X_full))^2) / sum((y_full - mean(y_full))^2), 4),
    round(summary(final_ols)$r.squared, 4),
    round(1 - sum((y_full - predict(final_rf, X_full))^2) / sum((y_full - mean(y_full))^2), 4)
  ),
  `In-Sample RMSE` = c(
    round(sqrt(mean((y_full - predict(final_tree, X_full))^2)), 2),
    round(sqrt(mean((y_full - predict(final_ols, X_full))^2)), 2),
    round(sqrt(mean((y_full - predict(final_rf, X_full))^2)), 2)
  ),
  `OOS R2` = c(
    round(mean(tree_r2_folds), 4),
    round(mean(ols_r2_folds), 4),
    round(mean(rf_r2_folds), 4)
  ),
  `OOS RMSE` = c(
    round(mean(tree_rmse_folds), 2),
    round(mean(ols_rmse_folds), 2),
    round(mean(rf_rmse_folds), 2)
  ),
  check.names = FALSE
)

kable(results,
      caption = "Model Performance: In-Sample vs Out-of-Sample",
      align = "lcccc") %>%
  kable_styling(full_width = FALSE)

```